Understanding the Data Duplication in VM Cloud Storage

### 2012-03-18 data_analysis/con.tex
update

\section{Conclusion}
In this paper, a empirical study using Aliyun's real user VM data traces is conducted. 
We present a number of observations and insights on the data duplication
in VM snapshot backups. We describe
in detail the patterns of VM access, VM snapshot duplication, locality reduction, and
the dintinct difference between OS and user generated data.
We hope this will facilitate future designs of cloud storage for VMs.

>>>>

\section{Conclusion}
In this paper, a empirical study using Aliyun's real user VM data traces is conducted. 
We present a number of observations and insights on where and how data duplication
exist in VM snapshot backups. We describe
in detail the patterns of VM access, the potential of deduplication, 
important factors such as locality, similarity and popularity, and
the dintinct difference between OS and user generated data duplication.
We hope this study will facilitate researchers and engineers to design better
data deduplication strategies
for backup storage in VM Cloud.

### 2012-03-18 data_analysis/dup.tex
update

\section{Heavily Duplicated Data}
Knowing that locality has ruled the 
data reduction within individual VM's snapshot backups,
in this section we investigate the dominant factor
of data duplication between VMs.

@@

We examine this by using the public VM images in Aliyun's cloud environment as a reference.
Every VM is generated from one of the public VM images by copying its data from base image,
so in order to see whether these data are changed after VM usage,
we first divide the public images into 4KB variable-sized blocks, then for each type of
OS, we check its 50 VM snapshot backups in VOSS, to see if this data block still exist in all
50 backups. This is a very strict criteria, a data block will only be marked as ``unchanged''
if it appear in all 50 snapshots.

\begin{figure}
  \centering
  %\epsfig{file=dedup_ratio.eps, width=3.2in}
  \includegraphics[width=3in,height=2in]{os_subset.png}
  \caption{}
  \label{fig:osdup}
\end{figure}

>>>>

\section{Heavily Duplicated Data}
\label{sect:dup}
Knowing that locality has ruled the 
data reduction within individual VM's snapshot backups,
in this section we investigate the dominant factor
of data duplication between VMs.

@@

We examine this by using the public VM images in Aliyun's cloud environment as a reference.
Every VM is generated from one of the public VM images by copying its data from base image,
so in order to see whether these data are changed after VM usage,
we first split the public images into 4KB variable-sized blocks, then for each type of
OS, we check its 50 VM snapshot backups in VOSS, to see if this data block still exist in all
50 backups. This is a very strict criteria, a data block will only be marked as ``unchanged''
if it appear in all 50 snapshots.

\begin{figure}
  \centering
  %\epsfig{file=dedup_ratio.eps, width=3.2in}
  \includegraphics[width=3in,height=2in]{os_subset.png}
  \caption{Data that almost never change in OS disks}
  \label{fig:osdup}
\end{figure}

### 2012-03-18 data_analysis/intro.tex
update

Unlike the legacy backup systems\cite{jumbo07} dealing with general file-level backup and deduplication, 
Backing up VM images is different. Although each VM image is treated as a file logically, its size is very large.
On the other hand, a cloud must support parallel backup of a large number of virtual disks everyday. 
One key requirement we face at Aliyun is that VM snapshot backup should only use a minimal amount of system
resources so that most of resources is kept for regular cloud system services or applications.
Thus our objective is to exploit the characteristics and duplication pattern of VM snapshot data and
pursue a cost-effective deduplication solution.

@@

In this paper we present the data analysis of Aliyun's VM snapshot data.
Our work is different from previous studies at several aspects:
First, we are targeting at the problem of snapshot backups, 
and no previous study has studied VM image with backup data involved.
Second, we focus on observing the pattern of data duplication in VM snapshot backups,
rather than examining the effect of variable-sized chunking algorithm.
Finally, we use real user's VM data rather than hand-made VM images.

\section{Experiment Setup}
We sampled two data sets from Aliyun's public VM cluster, all VMs
are used by real external users running various applications such like
database, web server, rendering services or even Hadoop. Each VM has
two virtual disks, one is for OS and software installations, and the other one
is for storing user data contents.

Data set VOSS composes of the OS disks from 35 VMs in 7 popular OSes: 
Debian, Ubuntu, Redhat, CentOS, Win2003 32bit, win2003 64 bit and win2008 64 bit. For each OS, 
5 VMs are chosen, and every VM come with 10 full snapshots of it OS disk. So
there is 350 full snapshot backups in this data set, overall size is about 7 TB.
We use VOSS to study the backup duplication characteristics, and OS disk change patterns.

Data set DDS contains the first snapshots of 1323 VMs' data disks from a cluster with 100+ nodes. 
Since no backup duplication is involved in this data set, this data set helps us to 
study the duplication pattern of user generated data. The overall size of dataset2 is near 23 TB.

>>>>

Unlike the legacy backup systems\cite{jumbo07} dealing with general file-level backup and deduplication, 
Backing up VM images is different: although each VM image is treated as a file logically from
external point of view, its size is very large.
On the other hand, a cloud must support parallel backup of a large number of virtual disks everyday. 
Two key requirements we face during desinging a backup storage system for VM snapshot are: 
\begin{enumerate}
\item VM snapshot backup should only use a minimal amount of system
resources so that most of resources is kept for regular cloud system services and VM themselves.
\item The entire snapshot storage and deduplication process must be fully decentralized to acheive
high scalibility and throughput, 
no component shall become a bottleneck.
\end{enumerate}

It is impossible to accomplish such requirements without fully utilizing the data duplication pattern
in VM snapshot backups
and design the optimal data deduplication strategies.
Thus we believe the first step towarding a cost-effective deduplication solution
 is to exploit the characteristics and duplication pattern of VM snapshot data.

@@

In this paper we present the data analysis of Aliyun's VM snapshot data,
our focus is to find out exploitable data duplication patterns and
correspoding factors.
Our work differs from previous studies at several aspects:
First, we are targeting at the problem of snapshot backups, 
and no previous study has studied VM image with backup data involved.
Second, we focus on observing the pattern of data duplication in VM snapshot backups,
rather than examining the effect of variable-sized chunking algorithm.
Finally, we use real user's VM data rather than hand-made VM images.

The rest of the paper is organized as follows: Section~\ref{sect:setup} introduces
the experiment setups, Section~\ref{sect:dedup} studies the potential of deduplication
in VM snapshot backups, Section~\ref{sect:loc} discusses the locality factor
in reduction of backup data,
Section~\ref{sect:scale} analysis the change of data duplication against system scale,
Section~\ref{sect:dup} introduces the patterns of heavily duplicated data.

\section{Experiment Setup}
\label{sect:setup}
We sampled two data sets from Aliyun's public VM cluster, where all VMs
are used by real world users running various applications such like
database, web server, rendering services or even Hadoop. Each VM has
two virtual disks, one is for OS and software installations, and the other one
is for storing user data contents.

Data set VOSS composes of the OS disks from 35 VMs in 7 popular OSes: 
Debian, Ubuntu, Redhat, CentOS, Win2003 32bit, win2003 64 bit and win2008 64 bit. For each OS, 
5 VMs are chosen, and every VM come with 10 full snapshots of it OS disk. So
there is 350 full snapshot backups in this data set, the overall size is about 7 TB.
We use VOSS to study the backup duplication characteristics and OS disk change patterns.

Data set DDS contains the first snapshots of 1323 VMs' data disks from a cluster with 100+ nodes. 
Since no backup duplication is involved in this data set, this data set helps us to 
study the duplication pattern of user generated data. The overall size of DDS is near 23 TB.

### 2012-03-18 data_analysis/locality.tex
update
initial version

\section{Impact of Locality}
Locality is widely used in many deduplication solutions, e.g., Zhu et al.~\cite{bottleneck08}
put in memory cache of block indices near the one which result in a disk index look-up and is found.
This is base on the observation that duplicates usually come in sequence rather than independently.

@@

\begin{figure*}[htb]
  \centering
  \includegraphics[width=6.5in,height=2in]{countmodify.png}
\caption{Bitmaps of dirty pages between snapshots}
\label{fig:dirty}
\end{figure*}

@@

>>>>

\section{Impact of Locality}
\label{sect:loc}
Locality is widely used in many deduplication solutions, e.g., Zhu et al.~\cite{bottleneck08}
put in memory cache of block indices near the one which result in a disk index look-up and is found.
This is base on the observation that duplicates usually come in sequence rather than independently.

@@

@@

\begin{figure*}[tbhp]
  \centering
  \includegraphics[width=6.5in,height=2in]{countmodify.png}
  %\epsfig{file=countmodify.eps, width=6.5in}
\caption{Bitmaps of dirty pages between snapshots}
\label{fig:dirty}
\end{figure*}

### 2012-03-18 data_analysis/overall.tex
update

\section{Overall Deduplication Effect}
Our study start from examining the effect of complete deduplication over both data sets.
Each virtual disk image is divided into small variable-sized blocks using the TTTD algorithm\cite{frame05},
under the average 4KB, maximum 16KB and minimum 2KB setting. Complete deduplication is
done by caculating the SHA-1 hash of each block and identifying duplicate copies base on compare-by-hash.

\begin{table*}[htb]
  \centering
    \begin{tabular}{|l|l|p{0.8in}|p{1.1in}|p{0.3in}|p{0.9in}|p{0.3in}|}
        \hline
        Data Set & OS Type & Original Size \newline (GB) & After Complete \newline Deduplication (GB) & Ratio & After 2MB Page \newline Reduction (GB) & Ratio \\ \hline
        \multirow{8}{*}{VOSS} & Debian & 1034.59 & 77.10 & 13.42 & 218.29 & 4.74 \\ \cline{2-7}
         & Ubuntu & 989.32 & 81.60 & 12.12 & 178.38 & 5.55 \\ \cline{2-7}
         & RHEL & 1007.28 & 62.70 & 16.07 & 215.33 & 4.68 \\ \cline{2-7}
         & CentOS & 973.03 & 57.34 & 16.97 & 522.53 & 1.86 \\ \cline{2-7}
         & Win2003 32Bit & 630.37 & 30.12 & 20.93 & 150.31 & 4.19 \\ \cline{2-7}
         & Win2003 64Bit & 793.47 & 44.16 & 17.97 & 167.15 & 4.75 \\ \cline{2-7}
         & Win2008 64Bit & 1508.97 & 46.39 & 32.52 & 222.54 & 6.78 \\ \cline{2-7}
         & Combined & 6937.03 & 389.65 & 17.80 & 1674.53 & 4.14 \\ \hline
        DDS & ~ & 23125.2 & 10887.2 & 2.12 & ~ & ~ \\
        \hline
    \end{tabular}
    \caption{Data reduction via complete deduplication and dirty page reduction}
    \label{tab:dedup}
\end{table*}

@@

\emph{Observation 1: Both methods reduce the data significantly, 
but there is still a big gap between the effect of dirty page method 
and what complete deduplication can acheive.} 
We believe complete deduplication can reduces 2x~5x more data than dirty page method for several resons:
First, the dirty page method doesn't resolve the duplication across VM backups. Second, the page size
is usually much bigger than actual range of modification, so many unchanged data are still backed up.

>>>>

\section{Overall Deduplication Effect}
\label{sect:dedup}

\begin{table*}[tbh]
  \centering
    \begin{tabular}{|l|l|p{0.8in}|p{1.1in}|p{0.3in}|p{0.9in}|p{0.3in}|}
        \hline
        Data Set & OS Type & Original Size \newline (GB) & After Complete \newline Deduplication (GB) & Ratio & After 2MB Page \newline Reduction (GB) & Ratio \\ \hline
        \multirow{8}{*}{VOSS} & Debian & 1034.59 & 77.10 & 13.42 & 218.29 & 4.74 \\ \cline{2-7}
         & Ubuntu & 989.32 & 81.60 & 12.12 & 178.38 & 5.55 \\ \cline{2-7}
         & RHEL & 1007.28 & 62.70 & 16.07 & 215.33 & 4.68 \\ \cline{2-7}
         & CentOS & 973.03 & 57.34 & 16.97 & 522.53 & 1.86 \\ \cline{2-7}
         & Win2003 32Bit & 630.37 & 30.12 & 20.93 & 150.31 & 4.19 \\ \cline{2-7}
         & Win2003 64Bit & 793.47 & 44.16 & 17.97 & 167.15 & 4.75 \\ \cline{2-7}
         & Win2008 64Bit & 1508.97 & 46.39 & 32.52 & 222.54 & 6.78 \\ \cline{2-7}
         & Combined & 6937.03 & 389.65 & 17.80 & 1674.53 & 4.14 \\ \hline
        DDS & ~ & 23125.2 & 10887.2 & 2.12 & ~ & ~ \\
        \hline
    \end{tabular}
    \caption{Data reduction via complete deduplication and dirty page reduction}
    \label{tab:dedup}
\end{table*}

Our study start from examining the effect of complete deduplication over both data sets.
Each virtual disk image is divided into small variable-sized blocks using the TTTD algorithm\cite{frame05},
under the average 4KB, maximum 16KB and minimum 2KB setting. Complete deduplication is
done by caculating the SHA-1 hash of each block and identifying duplicate copies base on compare-by-hash.

@@

\emph{Observation 1: Both methods reduce the data significantly, 
but there is still a big gap between the effect of dirty page method 
and what complete deduplication can accomplish.} 
We believe complete deduplication can reduces 2x~5x more data than dirty page method for several resons:
First, the dirty page method doesn't resolve the duplication across VM backups. Second, the page size
is usually much bigger than actual range of modification, so many unchanged data are still backed up.
